---
title: "Human Activity Recognition"
author: "Alexander Alexandrov"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
mode: selfcontained
---

## Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly.

This project goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.

The data for this project comes from this source: http://groupware.les.inf.puc-rio.br/har

## Exploratory Analysis

```{r, message=F, warning=F}
# Load necessary packages
library(data.table)
library(dplyr)
library(caret)
library(ggplot2)
library(randomForest)
```

Read training data.

```{r}
# Data table is used due to performance purposes
setwd(".")
training <- fread("pml-training.csv")
str(training)
```

* Some of variables contain NAs or empty strings. Such variables should be considered for interpolation or should be removed from further analysis.
* Window can be used to aggregate and reduce training data.

The approach is simple. Start from simplified model on reduced data. Check accuracy. Move to more complicated model if necessary.

\newpage

## Data Cleaning

Exclude columns with lots of NAs. Such columns can't be interpolated so can't be useful in machine learning.

```{r}
predictors.na.stat <- training[, colMeans(is.na(.SD) | .SD == "")]
# Less than 10% NAs
non.na.predictors <- names(training)[melt(predictors.na.stat) < 0.1]
```

Also predictors with nonrelevant information should be removed too.
User name and timestapms can't help to classify activity in common case.

```{r}
nonrelevant.predictors <- c("V1", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window")
relevant.predictors <- non.na.predictors[!(non.na.predictors %in% nonrelevant.predictors)]
relevant.training <- training[, relevant.predictors, with=FALSE]
```

## Further Exploratory Analysis

Just before averaging by window outliers should be considered.

```{r}
boxplot(relevant.training[, -c("num_window", "classe"), with=FALSE])
```

According to this plot a lot of predictors contains ourliers (black circles).
For example *Box-Cox* can be used to reduce outliers influence.

\newpage

## Final Data Cleaning

Average measurements by window to reduce original data set.

```{r}
cleaned.training <- relevant.training[, lapply(.SD, mean), by=c("num_window", "classe")]
# Exclude window variable
cleaned.training <- cleaned.training[, -"num_window", with=FALSE]
```

## Data Preprocessing

Exclude covariate predictors by means of correlation matrix

```{r}
cleaned.measurements <- cleaned.training[, -"classe", with=FALSE]
predictors.cor <- abs(cor(cleaned.measurements))
predictors.cor[upper.tri(predictors.cor, diag=TRUE)] <- 0

# Correlation threshold is 0.8
predictors.cor.coords <- which(predictors.cor > 0.8, arr.ind=TRUE)
predictors.cor.coords.x <- unique(predictors.cor.coords[, "col"])
predictors.cor.coords.y <- unique(predictors.cor.coords[, "row"])

covariate.predictors.indices <- if (length(predictors.cor.coords.x) > length(predictors.cor.coords.y)) {
  predictors.cor.coords.x
} else {
  predictors.cor.coords.y
}

covariate.predictors <- names(cleaned.measurements)[covariate.predictors.indices]

reduced.training <- cleaned.training[, -covariate.predictors, with=FALSE]
```

## Random Forest

This method is simple enough to get started and powerful to fit nonlinear case.

```{r}
set.seed(1234)
rf.model.fit <- train(classe ~ ., data=reduced.training, method="rf", ntree=500, trainControl="cv", number=25, preProcess="BoxCox")
rf.model.fit
```

## Conclusion

The accuracy is greater than *0.8*. So this model works practically fine. To avoid overfitting this model can be choosen as final.

## Testing

```{r}
testing <- fread("pml-testing.csv")
predicted <- predict(rf.model.fit, newdata=testing)
cbind(testing[, c("user_name", "cvtd_timestamp"), with=FALSE],
      data.frame(classe=predicted))
```
